Two applications of logic programming to Coq

### 2020-10-30 papers/types2020/app.tex
progress on 4

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  LJF

>>>>

\subsection{Semantics of the typed arithmetic language}
\label{ssec:types}

\begin{lstlisting}
Inductive has_type : tm -> typ -> Prop :=
   | T_Tru : has_type ttrue TBool
   | T_Fls : has_type tfalse TBool
    | T_Test : forall t1 t2 t3 T,
        has_type t1 TBool -> has_type t2 T -> has_type t3 T ->  has_type (tif  t1 t2 t3) T 
   | T_Zro : has_type tzero TNat
   | T_Scc : forall t1,
         has_type t1 TNat -> has_type (tsucc t1) TNat
   | T_Prd : forall t1,
        has_type t1 TNat -> has_type (tpred t1 ) TNat
   | T_Iszro : forall t1,
        has_type t1 TNat ->  has_type (tiszero t1) TBool.
Inductive nvalue : tm -> Prop :=
| nv_zero : nvalue tzero   | nv_succ : forall t, nvalue t -> nvalue (tsucc t).
Inductive bvalue : tm -> Prop :=
| bv_t : bvalue ttrue   | bv_f : bvalue tfalse. 
 Reserved Notation "t1 '===>' t2" (at level 40).
 Inductive step : tm -> tm -> Prop :=
   | ST_IfTrue : forall t1 t2,
       (tif ttrue t1 t2) ===> t1
   | ST_IfFalse : forall t1 t2,
       (tif tfalse t1 t2) ===> t2
   | ST_If : forall t1 t1' t2 t3,
       t1 ===> t1' -> (tif t1 t2 t3) ===> (tif t1' t2 t3)
   | ST_Succ : forall t1 t1',
       t1 ===> t1' -> (tsucc t1) ===> (tsucc t1')
   | ST_PredZero : (tpred tzero) ===> tzero
   | ST_PredSucc : forall t1,
       nvalue t1 -> (tpred (tsucc t1)) ===> t1
   | ST_Pred : forall t1 t1',
       t1 ===> t1' -> (tpred t1) ===> (tpred t1')
   | ST_IszeroZero : (tiszero tzero) ===> ttrue
   | ST_IszeroSucc : forall t1,
        nvalue t1 ->  (tiszero (tsucc t1)) ===> tfalse
   | ST_Iszero : forall t1 t1',
       t1 ===> t1' -> (tiszero t1) ===> (tiszero t1')
 where "t1 '===>' t2" := (step t1 t2).
Inductive notstuck (e : tm) (Step : tm -> tm -> Prop) : Prop :=
  | pn : nvalue e  -> notstuck e Step
  | pb : bvalue e -> notstuck e Step
  | ps e' : Step e e' -> notstuck e Step.
\end{lstlisting}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  LJF Interp dep pbt coq elpi

### 2020-10-30 papers/types2020/conc.tex
progress on 4

\begin{itemize}
\item Brief discussion/comparison with \texttt{auto}
\item Mention other papers by Enrico as applications of coq-elpi
\item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in say MetaCoq would be a disaster (reimplement unfication, backtracking, binders)
\end{itemize}

@@

A mention to the use of \emph{delays} in \texttt{check}: instead of generating only ground terms, we could generate say lists with at least two elements as \texttt{[1;2|Tail]} or simply not commit to a base type. This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or the stuff by Fabio Fioravanti).

>>>>

\begin{itemize}
\item Brief discussion/comparison with \texttt{auto}
\item Mention other papers by Enrico as applications of coq-elpi~\cite{coen19mscs,tassi19itp}\dots
\item Say we \textbf{do} not talk about Coq metaprogramming (Ltac, Mtac), do say that doing our to application in MetaCoq~\cite{sozeau2020metacoq} would be a disaster (reimplement unfication, backtracking, binders)
\end{itemize}

@@

A mention to the use of \emph{delays} in \texttt{check}: instead of generating only ground terms, we could generate say lists with at least two elements as \texttt{[1;2|Tail]} or simply not commit to a base type. This  is similar to the use of needed narrowing in LazySmallCheck to generate partial trace or the idea of using CLP for data generation (see Dubois' FocalTest or~\cite{FioravantiPS15}).

### 2020-10-30 papers/types2020/main.tex
progress on 4

\begin{abstract}
The \emph{foundational proof certificate} (FPC) framework is capable
of checking and elaborating a wide range of proof evidence and can easily
be implemented using the logic programming language \lP.
%
% The checking of a proof certificate in this higher-order logic
% programming language can yield a fully detailed, dependently
% typed $\lambda$-terms that can be checked by the Coq kernel.
%
We use the recently released Coq-Elpi plugin for Coq in which the Elpi
implementation of \lP can send and retrive information to and from the Coq kernel.
%
We provide two examples: first we implement a FPC-driven sequent calculus for a fragment of the Calculus of Inductive Constructions and we package it into a tactic to perform property-based testing of Inductive types corresponding to pure (higher-order) Horn programs. Secondly, we implement a proof checker for first-order LJF to demonstrate how  proof certificates
can be supplied by external (to Coq) provers and elaborated into the
fully detailed proof terms that can be checked by the Coq kernel.
\end{abstract}

@@

\input conc
\bibliography{l}
\appendix
\input app
\end{document}

>>>>

\begin{abstract}
The \emph{foundational proof certificate} (FPC) framework is capable
of checking and elaborating a wide range of proof evidence and can easily
be implemented using the logic programming language \lP.
We use the recently released Coq-Elpi plugin for Coq in which the Elpi
implementation of \lP can send and retrieve information to and from the Coq kernel.
We provide two examples: first we implement a FPC-driven sequent calculus for a fragment of the Calculus of Inductive Constructions and we package it into a tactic to perform property-based testing of Inductive types corresponding to pure (higher-order) Horn programs. Secondly, we implement a proof checker for first-order LJF to demonstrate how  proof certificates
can be supplied by external (to Coq) provers and elaborated into the
fully detailed proof terms that can be checked by the Coq kernel.
\end{abstract}

@@

\input conc
\bibliography{l}
\newpage
\appendix
\input app
\end{document}

### 2020-10-30 papers/types2020/pbt.tex
progress on 4

\todo{the \texttt{aexp} example}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO

>>>>

%\todo{the \texttt{aexp} example}

For a more significant case study, let us turn to the semantic of programming languages, where PBT has been used extensively and successfully~\cite{Klein12}. Here we will use a far simple example, the a typed arithmetic language featuring numerals with predecessor and Booleans with if-then-else and test-for-zero, which comes from (\url{https://softwarefoundations.cis.upenn.edu/plf-current/Types.html}):
\begin{lstlisting}
Inductive tm : Type :=
 | ttrue : tm | tfalse : tm | tif : tm -> tm -> tm -> tm | tzero : tm | tsucc : tm -> tm
 | tpred : tm -> tm | tiszero : tm -> tm.
 Inductive typ : Type := |TBool : typ |TNat : typ.
\end{lstlisting}
The static and small step dynamic semantics are reported in appendix~\ref{ssec:types}.

While it is obvious that \emph{subject expansion} fails for this calculus, it is gratifying to have it confirmed, with \lsti{e = tif ttrue tzero ttrue}:
\begin{lstlisting}
Conjecture subexp:  forall e e' t, step e e' -> has_type e' t -> has_type e t. 
 intros e e' t HS HT.
 elpi dep_pbt height 2 (HS /\ HT) (e).
\end{lstlisting}

Another way to asses the fault detection capability of our PBT setup is via \emph{mutation analysis}, whereby localized bugs are voluntarily inserted, with the view that they should caught (``killed'') by the testing suite. Following on  an exercise if the aforementioned chapter of SF, we modify the typing relation by adding the following additional clause:

\begin{lstlisting}
Module M1.
Inductive has_type : tm -> typ -> Prop :=
$\dots$
  | T_SuccBool : forall t, has_type t TBool -> has_type (tsucc t) TBool.
end M1.
\end{lstlisting}
Some of the required properties for a typing system now fails, not
only type uniqueness, but also progress with counterexample \lsti{e = tsucc ttrue}:
\begin{lstlisting}
Definition progress (e :tm) (Has_type : tm -> typ -> Prop) (Step : tm -> tm -> Prop):= 
     forall t, Has_type e t -> notstuck e Step.
Conjecture progress_m1:  forall e, progress e M1.has_type step.
 unfold progress.
 intros e t Ht.    
 elpi dep_pbt 2 height (Ht) (e) .  
\end{lstlisting}
Whereas the \emph{Types} examples is admittedly quite simple-minded,
it has been used as a benchmark for evaluating \textsf{QuickChick}'s automation of generators under invariants~\cite{LampropoulosPP18}, and to be amenable to the tool, the specification had to be massaged non-insignificantly. 
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  Bedwyr bipole elpi QuickCheck OCaml TODO coq xs nat
% LocalWords:  tif ttrue tzero tsucc

### 2020-10-30 papers/types2020/sec3.tex
progress on 4

Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
%
\todo{Explain relations with other term assignments. Note that we do not use explicit substitutions (we are cut free). Explain the choice of not having certificates represent proof terms. Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality: this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that are inherited by Coq }

Figure~\ref{fig:kernel} contains the elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate.
%
% Notice that it is easy to show that no matter how the expert predicates
% are defined, if the goal \lsti{check Cert B} is provable in \lP then
% \lsti{B} must be a sound consequence of the program clauses stored in
% \dots
% the \lsti{prog} predicate (which provides a natural implementation of
% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

@@

\begin{lstlisting}
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
}}.
\end{lstlisting}
The following example show how we can do FPC-driven logic programming in Coq and return
a Coq term
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog 10.
Qed.
Print i1.
\end{lstlisting}
\begin{verbatim}
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)))
	 : exists R : list nat, insert 2 [0; 1] R
\end{verbatim}
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

>>>>

Figure~\ref{fig:augmented} contains a simple proof system \dots  in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
%
\todo{Explain relations with other term assignments. Note that we do not use explicit substitutions (we are cut free). Explain the choice of not having certificates represent proof terms. Explain that the unfold clause handles defined atoms and defined connectives (conjunction, disjunction, existential and also equality): this may be less efficient and proof-theoretically induce pointless phase changes, but it is arguably very elegant e simplifies proof terms that arand inherited by Coq }

Figure~\ref{fig:kernel} contains the Elpi implementation of the
inference rules in Figure~\ref{fig:augmented}: here the infix
turnstile $\vdash$ symbol is replaced by the \lsti{check} predicate.
%
% Notice that it is easy to show that no matter how the expert predicates
% are defined, if the goal \lsti{check Cert B} is provable in \lP then
% \lsti{B} must be a sound consequence of the program clauses stored in
% \dots
% the \lsti{prog} predicate (which provides a natural implementation of
% the premise $(A~\hbox{\tt :-}~G)\in\instan{\Pscr}$.
% AM: not there
% in Figure~\ref{fig:kernel}).

@@

\begin{lstlisting}
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
}}.
\end{lstlisting}
The following example show how we can do FPC-driven logic programming in Coq and return
a Coq term
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}

%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
